\documentclass[]{article}
\usepackage[margin=1.0in]{geometry}
\usepackage{amssymb}

%title material
\title{Astronomy 400B Lecture 5: Stellar Orbits}
\author{Brant Robertson}
\date{February, 2015}


%include latex definitions
\input{astro400B_definitions.tex}

%begin the document
\begin{document}

%make the title, goes after document begins
\maketitle

%first section
\section{Motion Under Gravity}

Newton's Law of Gravity point mass $M$ attracts another mass $m$ separated by 
distance $\vr$, causing a change in momentum $m\vv$ of:

\begin{equation}
\frac{d}{dt}(m\vv) = -\frac{GmM}{r^{3}} \vr
\end{equation}
\noindent
where $G$ is Newton's gravitational constant. For an $N$-body system, we have
\begin{equation}
\frac{d}{dt}(m_{i}\vv_{i}) = -\sum_{j\ne i}\frac{G m_{i} m_{j}}{|\vx_i - \vx_j|^{3}} (\vx_i - \vx_j)
\end{equation}
\noindent
This equation can be re-written as
\begin{equation}
\frac{d}{dt}(m \vv_i) = - m \nabla\Phi(\vx_i)
\end{equation}
\noindent
where
\begin{equation}
\Phi(\vx_i) = - \sum_{i\ne j} \frac{G m_j}{|\vx_i - \vx_j|}
\end{equation}
\noindent
is the gravitational potential supplied by the point mass distribution at positions $\vx_i$.
Note we have chosen to define the potential such that $\Phi(x)\to0$ as $x\to\infty$ but
this is arbitrary. Note that
\begin{equation}
\nabla = \left[\frac{\partial}{\partial x},\frac{\partial}{\partial y},\frac{\partial}{\partial x}\right]
\end{equation}

\subsection{Continuous Matter Distributions}
Now consider a continuous distribution of matter density $\rho(\vx)$.  The
potential generated by $\rho(\vx)$ is given by
\begin{equation}
\label{eqn:potential}
\Phi(\vx) = - \int \frac{G\rho(\vx')}{|\vx - \vx'|}d^{3}\vx'
\end{equation}
\noindent
Note that the integral is performed over $\vx'$.  The force $\vF$ per unit mass
is
\begin{equation}
\vF(\vx) = - \nabla \Phi(\vx) = - \int \frac{G\rho(\vx')(\vx-\vx')}{|\vx-\vx'|^{3}}d^{3}\vx'
\end{equation}

\subsection{Poisson's Equation}

Take Equation \ref{eqn:potential} and apply the Laplacian operator 
\begin{equation}
\nabla^{2} \equiv \nabla \cdot \nabla = \left[\frac{\partial^{2}}{\partial x^{2}}+\frac{\partial^{2}}{\partial y^{2}}+\frac{\partial^{2}}{\partial z^{2}}\right]
\end{equation}
\noindent
to both sides.  Remembering that the operator acts on $\vx$ and not $\vx'$, we have
\begin{equation}
\label{eqn:pois_init}
\nabla^{2} \Phi(\vx) = - \int G \rho(\vx') \nabla^{2} \left(\frac{1}{|\vx-\vx'|}\right)d^{3}\vx'.
\end{equation}
\noindent
We can evaluate this by noting that
\begin{equation}
\label{eqn:nabla}
\nabla\left(\frac{1}{|\vx - \vx'|}\right) = -\frac{\vx-\vx'}{|\vx-\vx'|^3}, \nabla^{2}\left(\frac{1}{|\vx - \vx'|}\right) = 0.
\end{equation}
\noindent
So we conclude that outside of a very small region around $\vx$, $\nabla^{2}\Phi(\vx)=0$.
Let's take a spherical region $S(\epsilon)$ of radius $\epsilon$ centered on $\vx$. In proceeding, let's 
note that 
\begin{equation}
\nabla^{2} f(|\vx-\vx'|) = \nabla_{\vx'}^{2} f(|\vx-\vx'|)
\end{equation}
\noindent
for any function $f(|\vx - \vx'|)$. If we take $\epsilon$ to be small enough such 
that $\rho(\vx)\approx$ a constant, then we can write
\begin{eqnarray}
\label{eqn:pois_med}
\nabla^{2}\Phi(\vx) &\approx& - G\rho(\vx) \int_{S(\epsilon)} \nabla^2 \left( \frac{1}{|\vx - \vx'|} \right)d^{3}\vx' \nonumber \\
&=& - G \rho(\vx)  \int_{S(\epsilon)} \nabla_{\vx'}^2 \left( \frac{1}{|\vx - \vx'|} \right)d V'.
\end{eqnarray}
\noindent
Now we get to use the {\it divergence} theorem
\begin{equation}
\int \nabla^{2} f dV = \oint \nabla f \cdot dS,
\end{equation}
\noindent
which allows us to write Equation \ref{eqn:pois_med} as
\begin{equation}
- G \rho(\vx)  \int_{S(\epsilon)} \nabla_{\vx'}^2 \left( \frac{1}{|\vx - \vx'|} \right)d V' = - G \rho(\vx) \oint_{S(\epsilon)} \nabla_{\vx'}\left(\frac{1}{|\vx-\vx'|}\right) \cdot d\vS'
\end{equation}
\noindent
By applying Equation \ref{eqn:nabla} and the identity $\nabla_{\vx'} f = -\nabla f$, we have
\begin{eqnarray}
- G \rho(\vx) \oint_{S(\epsilon)} \nabla_{\vx'}\left(\frac{1}{|\vx-\vx'|}\right) \cdot d\vS' &=&
-G \rho(\vx) \oint_{S(\epsilon)} \left(\frac{\vx-\vx'}{|\vx-\vx'|^{3}}\right) \cdot d\vS' \nonumber \\
&=& 4 \pi G \rho(\vx)
\end{eqnarray}


\subsection{Inside a Uniform Shell}

The gravitational force inside a spherical shell of uniform density is zero.  The potential is a
constant.

{\bf See Figure 3.1 of Sparke and Gallagher.}

The opening angle OA is the same as OB, so the ratio of the enclosed mass is (SA/SB)$^{2}$.  Since
the ratio of the forces scale like the inverse of this ratio (from the inverse square law), the
force contributions of the A and B patches are equal and opposite.

\subsection{Gravitational Potential Outside a Uniform Spherical Shell}

{\bf See Figure 3.2 of Sparke and Gallagher}

We are calculating the potential a uniform spherical shell of mass $M$ and radius $a$.  
Consider a point $P$ a distance $r$. The contribution of a narrow cone of opening
solid angle $\Delta\Omega$ around another point $Q'$ is
\begin{equation}
\Delta\Phi[\vx(P)] = -\frac{GM}{|\vx(P) - \vx(Q')|} \frac{\Delta\Omega}{4\pi}
\end{equation}
\noindent
Now consider the potential $\Phi'$ at point $P'$ at a radius $a$ away from the
center of a shell of the same mass $M$ but with a radius $r$.
The contribution $\Delta\Phi'$ from the material in the same cone of solid
angle $\Delta\Omega$ but at point $Q$ a distance $r$ away is
\begin{equation}
\Delta\Phi'[\vx(P')] = - \frac{GM}{|\vx(P') - \vx(Q)|}\frac{\Delta\Omega}{4\pi}
\end{equation}
\noindent
but since $PQ'=P'Q$, $\Delta\Phi[\vx(P)] = \Delta\Phi'[\vx(P')]$.
When we integrate over $4\pi$, we have
\begin{equation}
\Phi[\vx(P)] = \Phi'[\vx(P')]= \Phi'[\vx=0] = -\frac{GM}{r}
\end{equation}
\noindent
The force associated with this spherical shell is just $F(r) = \nabla \Phi[\vx(P)] = -\frac{GMm}{r}$.
So the force outside the shell is the same as for a point mass at distance $r$.

Inside a spherical mass distribution $\rho(r)$, the 
centripetal acceleration that allows for a circular 
orbit must be the radial gravitational force inwards.
On a circular orbit, in terms of the circular velocity
$V$ this acceleration is just
\begin{equation}
a = \frac{V^{2}(r)}{r} = -F(r) = \frac{GM(<r)}{r^2}.
\end{equation}
\noindent
For a point mass, $V(r)\propto r^{-1/2}$.  No extended
distribution can have a circular velocity curve that
declines more rapidly than $\propto r^{-1/2}$.

Note that the potential of a distributed mass density $\rho(\vx)$
is not the same as for a point mass.  Instead, we have
\begin{equation}
\Phi(r) = - \left[ \frac{GM(<r)}{r} + 4 \pi G \int_{r}^{\infty} \rho(r') r' dr'\right].
\end{equation}
\noindent
But as long as the spherical mass distribution has a finite size, eventually we 
will have
\begin{equation}
\Phi(\vx) \to - \frac{G M_{\mathrm{tot}}}{|\vx|}
\end{equation}
\noindent
at large enough radius.


\subsection{Moving Through a Potential}

If we are moving through a background potential $\Phi(\vx)$ with
velocity $\vx$, the potential we experience changes
with time according to $d\Phi/dt = \vv\cdot\nabla\Phi(\vx)$.
We can re-write Newton's equation as
\begin{equation}
\vv \cdot \frac{d}{dt} ( m\vv) + m \vv\cdot\nabla \Phi(\vx) = 0 = \frac{d}{dt}\left[\frac{1}{2}m\vv^{2} + m \Phi(\vx)\right]
\end{equation}
\noindent
Therefore, the total energy
\begin{equation}
\label{eqn:total_energy}
E \equiv \frac{1}{2} m\vv^{2} + m\Phi(\vx) =~\mathrm{const}
\end{equation}
\noindent.
We can write of course that $E = KE + PE$, where $KE = \frac{1}{2}m\vv^{2}$ and $PE = m\Phi(\vx)$.
The kinetic energy cannot be negative, and we adopt $\Phi(\vx)\to0$ as $\vx\to\infty$.  At position
$\vx$, an orbit is unbound only if the total energy $E>0$. The speed at this place in the
orbit must exceed the escape speed, which is found by setting Equation \ref{eqn:total_energy} to zero.
We then have
\begin{equation}
v_{e}^{2} = - 2 \Phi(\vx).
\end{equation}

\subsection{Angular Momentum}

The angular momentum of an orbit is $L = \vx \times m\vv$. The time rate of change is
\begin{equation}
\frac{dL}{dt} = \vx \times \frac{d}{dt} (m \vv) = -m\vx \times \nabla \Phi.
\end{equation}
\noindent
For a spherically symmetric distribution, the force is central and $dL/dt = 0$ (angular
momentum is conserved).
In an axisymmetric distribution, on the component of $L$ parallel to the symmetry
axis is conserved.

\subsection{Individual Energy is Not Conserved in a Time-Dependent Potential}

In a many-body system, the total energy of each star is not individually conserved.
The time derivative of the kinetic energy of star $i$ is
\begin{equation}
\sum_{i} \vv_i \cdot \frac{d}{dt} (m_i \vv_i) = \frac{d}{dt} KE = -\sum_{i,j;i\ne j} \frac{Gm_i m_j}{|\vx_i - \vx_j|^{3}}(\vx_i - \vx_j) \cdot \vv_i
\end{equation}
\noindent
Doing the same calculation on star $j$ and taking the dot product with $\vv_j$ gives
\begin{equation}
\frac{1}{2}\sum_{j} \frac{d}{dt} (m_j \vv_j \cdot \vv_j) = - \sum_{i,j; i\ne j}\frac{Gm_i m_j}{|\vx_i - \vx_j|^{3}}(\vx_j - \vx_i)\cdot \vv_j
\end{equation}
\noindent
Adding the RHS of these two equations gives
\begin{equation}
- \sum_{i,j; i\ne j}\frac{Gm_i m_j}{|\vx_i - \vx_j|^{3}}(\vx_i - \vx_j)\cdot(\vv_i - \vv_j) = \sum_{i,j;i\ne j}\frac{d}{dt}\left(\frac{Gm_i m_j}{|\vx_i - \vx_j|}\right).
\end{equation}
The potential energy $PE$ is a sum of pairs of potentials from individual objects
\begin{equation}
PE = -\frac{1}{2} \sum_{i,j;i\ne j}\frac{Gm_i m_j}{|\vx_i - \vx_j|} = \frac{1}{2} \sum_i m_i \Phi(\vx_i) = \frac{1}{2} \int \rho(\vx)\Phi(\vx)dV.
\end{equation}
\noindent
We divided by two so every object contributes only once.

We can now see that, for the whole collection of objects 
\begin{equation}
2 \frac{d}{dt} \left[ KE - \frac{1}{2} \sum_{i,j;i\ne j} \frac{G m_i m_j}{|\vx_i - \vx_j|}\right] = 0.
\end{equation}
\noindent
This means the total energy of the system is conserved.

\subsection{External Forces}

Consider the total force on an object $i$ in 
a many body system under the influence of an external force $\vF_{\mathrm{ext}}$.
\begin{equation}
\sum_i \frac{d}{dt} (m_i \vv_i) \cdot \vx_i = -\sum_{i,j; i\ne j} \frac{Gm_i m_j}{|\vx_i - \vx_j|^{3}} (\vx_i - \vx_j) \cdot \vx_i + \sum_i \vF_{\mathrm{ext}}^{i} \cdot \vx_i.
\end{equation}
\noindent
The force on the $j$th object is
\begin{equation}
\sum_j \frac{d}{dt} (m_j \vv_j) \cdot \vx_j = - \sum_{i,j;i\ne j }\frac{G m_i m_j}{|\vx_i - \vx_j|^{3}}(\vx_j - \vx_i)\cdot\vx_j + \sum_j \vF_{\mathrm{ext}}^{j} \cdot \vx_j
\end{equation}
\noindent
The left hand sides of these equations are equal, and are equal to
\begin{equation}
\frac{1}{2} \sum_i \frac{d^2}{dt^2}(m_i\vx_i \cdot \vx_i) - \sum_i m_i \vv_i \cdot \vv_i = \frac{1}{2} \frac{d^2I}{dt^2} -2 KE
\end{equation}
\noindent
where the moment of inertia $I$ is
\begin{equation}
I \equiv \sum_i m_i \vx_i \cdot \vx_i.
\end{equation}
\noindent
By averaging the force on $i$ and $j$, we find
\begin{equation}
\frac{1}{2} \frac{d^2 I}{dt^2} - 2 KE = PE + \sum_i \vF_{\mathrm{ext}}^{i} \cdot \vx_i
\end{equation}
\noindent
and averaging this over a short time interval $0<t<\tau$ gives
\begin{equation}
\frac{1}{2\tau}\left[ \frac{dI}{dt}(\tau) - \frac{dI}{dt}(0)\right] = 2 \ave{KE} + \ave{PE} + \sum_i \ave{\vF_{\mathrm{ext}}^{i} \cdot \vx_i}
\end{equation}
\noindent
If all objects in the system are bound, then $|\vx_i \cdot \vv_i|$ and $dI/dt$ will be finite.  As $\tau\to\infty$, the LHS goes to zero.  Then we have
\begin{equation}
2\ave{KE} + \ave{PE} + \sum_i \ave{\vF_{\mathrm{ext}}^{i} \cdot \vx_i} = 0
\end{equation}

\section{Two-Body Relaxation}

A potential can be thought of as a combination of a smooth and steep potential wells.
We can calculate the average time between strong encounters with the steep
potential wells supplied near individual stars. Suppose that the
stars have mass $m$ and typical velocities $V$. If two stars come
within $r$ of one another, their kinetic energies increase to balance the
change in potential energy.  We say there is a {\it strong encounter}
if the change in potential energy is comparable to their starting
kinetic energy.  In other words
\begin{equation}
\frac{Gm^{2}}{r} \gtrsim \frac{mV^{2}}{2}
\end{equation}
\noindent
and the radius must be smaller than
\begin{equation}
r \lesssim r_{s} \equiv \frac{2Gm}{V^2}.
\end{equation}
\noindent
In the solar neighborhood, $V\approx30~\km~s^{-1}$, and taking $m\sim0.5\Msun$ we have $r_s \approx 1~$AU.
For the Sun, there has been no strong encounter for $\sim4.5~\Gyr$. Over a time $t$, the Sun
has an encounter with all stars in a cylinder with volume $\pi r_s^2 V t$. If there is a number
density of $n$, then we are interested in the time when $n\pi r_s^2 V t = 1$.  This time is
\begin{equation}
t_s = \frac{V^3}{4\pi G^2 m^3 n} \approx 4 \times 10^{12}~\yr~\left(\frac{V}{10~\km~s^{-1}}\right)^3\left(\frac{m}{\Msun}\right)^{-2}\left(\frac{n}{1~\pc^{-3}}\right)^{-1}.
\end{equation}
\noindent
For $n\approx0.1~\pc^{-3}$, then $t_s\sim10^{15}~\yr$.


\subsection{Weak Encounters}

Instead of strong encounters, we need to calculate the effects of weaker encounters
on indivudal stars.  We do this via the impulse approximation, that tells us how to
approximate the change in velocity from a weak encounter.  As one object passes
another, the perpendicular force between them is
\begin{equation}
\vF_{\perp} = \frac{GmMb}{(b^2 + V^2t^2)^{3/2}} = M \frac{dV_{\perp}}{dt}.
\end{equation}
\noindent
Integrating over time, we find that
\begin{equation}
\Delta V_{\perp} = \frac{1}{M} \int_{-\infty}^{\infty} \vF_{\perp}(t) dt =\frac{2 G m}{bV}.
\end{equation}
Slower approaches result in larger perpendicular velocity changes. The
path of $M$ is bent via
\begin{equation}
\alpha = \frac{\Delta V_{\perp}}{V} = \frac{2Gm}{bV^2}.
\end{equation}
\noindent
A weak encounter requires $b$ to be larger than $r_s$.

The number of stars with mass $m$ passing $M$ with
separations between $b$ and $b+\Delta b$ is the product of
the number density $n$ and the volume $V t \cdot 2 \pi b\Delta b$.
We multiply by $\Delta V_{\perp}^{2}$ and integrate over $b$ to find
\begin{equation}
\ave{\Delta V_{\perp}^{2}} = \int_{b_{min}}^{b_{max}} n V t\left(\frac{2Gm}{bV}\right)^2 2 \pi b db = \frac{8 \pi G^2 m^2 nt}{V} \ln \left(\frac{b_{max}}{b_{min}}\right).
\end{equation}
\noindent
After a time $t_{\mathrm{relax}}$ such that $\ave{\Delta V_{\perp}^2} = V^2$, the expected perpendicular 
velocity becomes about equal to its original forward speed.  The initial path is forgotten!
Defining $\Lambda\equiv (b_{max}/b_{min})$, we find the relaxation time is much shorter than
the strong interaction time as
\begin{equation}
t_{\mathrm{relax}} = \frac{V^3}{8\pi G^2 m^2 n \ln \Lambda} = \frac{t_s}{2 \ln \Lambda}
\end{equation}
\begin{equation}
t_{\mathrm{relax}} \approx \frac{2\times10^9~\yr}{\ln \Lambda}\left(\frac{V}{10~\km~s^{-1}}\right)^3\left(\frac{m}{\Msun}\right)^{-2}\left(\frac{n}{10~\pc^{-3}}\right)^{-1}.
\end{equation}
\noindent
So what is $\Lambda$?  Well, if $b<r_s$ the method can't be correct and we usually take $b_{min} = r_s$.  We
can then take $b_{max}$ to be the size of the whole system.  For the Sun, $r_s = 1$~AU.  If we 
take $300~\pc \leq b_{max} \leq 30~\kpc$, then $\Lambda\approx18-22$.  So the exact value of
$\Lambda$ doesn't matter, as $t_{\mathrm{relax}}\approx10^{13}~\yr$.

In a cluster consisting of $N$ stars with mass $m$ and typical $V$, the average separation is about
half the size of the system $R$.  We have
\begin{equation}
\frac{1}{2}NmV^2 \sim \frac{G(Nm)^2}{2R},~\Lambda = \frac{R}{r_s} \sim \frac{GmN}{V^2} \cdot \frac{V^2}{2Gm}\sim\frac{N}{2}.
\end{equation}
The crossing time of the system is $t_{cross} \sim R/V$.  Since $N=4n\pi R^3 /3$, we have
\begin{equation}
\frac{t_{relax}}{t_{cross}} \sim \frac{V^4 R^2}{6 N G^2 m^2 \ln \Lambda} \sim \frac{N}{6 \ln (N/2)}.
\end{equation}
\noindent
For $N\sim10^{11}$ stars, the relaxation will take $10^9$ crossing times.  For globulars with $10^6$ stars,
the relaxation time for the whole system is $t_{\mathrm{relax}} \sim 10^4 t_{cross}$.

\subsection{Effects of Two-Body Relaxation}

Consider a cluster of stars, each with mass $m$.
The {\it Maxwellian} distribution gives the fraction $f$ of stars with 
velocities between $v$ and $v+\Delta v$ as
\begin{equation}
4\pi f_M(E) v^{2} \Delta v 
\end{equation}
\noindent
where
\begin{equation}
f_M(E)\propto \exp\left(-\frac{E}{kT} \right) = \exp \left\{ -\left[ m\Phi(\vx) + \frac{mv^{2}}{2} \right]
 \Big/ kT \right\}.
\end{equation}
The temperature is related to the average kinetic energy as
\begin{equation}
\frac{1}{2}m\ave{\vv^{2}(\vx)} = \frac{3}{2}kT.
\end{equation}
\noindent
But there are stars at large energies, and these will evaporate. The kinetic
energy of stars that escape will typically be
\begin{equation}
\left\langle \frac{1}{2} m v_2^2(\vx) \right\rangle = - \frac{1}{N} \sum_i m_i \Phi(\vx_i) = -\frac{2}{N} PE = \frac{4}{N} KE = 6 kT
\end{equation}
The fraction of stars that escape will be
\begin{equation}
\int_{\sqrt{12kT/m}}^{\infty} f(E) v^2 dv \Big/ \int_0^{\infty} f(E) v^2 dv = 0.0074 \approx \frac{1}{136}
\end{equation}
\noindent
The evaporation time will be $t_{evap} \sim 136 t_{relax}$.  For globulars this is longer than
the age of the universe, but for open clusters it is a few gigayears.

\section{Orbits of Disk Stars and Epicycles}

Now we will treat orbits in axisymmetric potentials, and we will use cylindrical polar
coordinates $(R,\phi,z)$. In an axisymmetric potential, $\partial\Phi/\partial\phi = 0$, and
angular momentum about the $z$-axis is conserved.  We adopt $l_z$ at the angular
momentum per unit mass, we can write
\begin{equation}
\frac{d}{dt}(R^2 \dot{\phi}) = 0;~l_z \equiv R^2 \dot{\phi} =~\mathrm{constant}.
\end{equation}
\noindent
The radial equation of motion is
\begin{equation}
\label{eqn:radial_motion}
\ddot{R} = R \dot{\phi}^2 - \frac{\partial \Phi}{\partial R} = - \frac{\partial \Phi_{\mathrm{eff}}}{\partial R}
\end{equation}
\noindent
\begin{equation}
\Phieff \equiv \Phi(R,z) + \frac{l_z^2}{2R^2}.
\end{equation}
Multiplying \ref{eqn:radial_motion} by $\dot{R}$ and integrating informs us that
for a star in the midplane $z=0$
\begin{equation}
\frac{1}{2}\dot{R}^2 + \Phieff(R,z=0;l_z) = ~\mathrm{constant}.
\end{equation}
\noindent
If $l_z\ne0$, then there is an angular momentum barrier at the center of the galaxy that
limits $R$ to some minimum value.

The vertical equation of motion is
\begin{equation}
\ddot{z} = -\frac{\partial \Phi}{\partial z} (R,z) = -\frac{\partial \Phieff}{\partial z} (R,z)
\end{equation}
\noindent
If the potential is an even function about $z=0$ there is no vertical force in the midplane.

Let's expand the potential about the average radius $R_g$ of a star in a Taylor series and
truncate at the first term
\begin{equation}
\ddot{z} = -z \left[ \frac{\partial^2 \Phi}{\partial z^2}(R_g, z)\right]_{z=0} \equiv - \nu^2(R_g) z
\end{equation}
\noindent
where $\nu$ is an angular frequency.  The solution to this equation is
a harmonic oscillator
\begin{equation}
z(t) = Z \cos (\nu t + \theta)
\end{equation}
\noindent
where $Z$ and $\theta$ are constants.

Circular orbits with $\dot{R} = 0$ are only possible at a radius $R_g$ where $\Phieff =~\mathrm{constant}$.
At this radius
\begin{equation}
\frac{\partial \Phi}{\partial R}(R_g,z=0) = \frac{l_z^2}{R_g^3} = R_g \Omega^{2}(R_g)
\end{equation}
\noindent
where $\Omega(R)$ is the angular speed of a circular orbit in the disk plane.  If $\Phieff$ has a
minimum at $R_g$, then the circular orbit will have the lowest $l_z$.  Any larger $l_z$
will oscillate about the circular orbit, in an {\it epicycle} about a point that
moves with $\Omega(R_g)$ in a circular orbit of radius $R_g$.

Let's set $R = R_g+x$ in Equation \ref{eqn:radial_motion}, assume that $x\ll R$, and expand to
first order in $x/R$ and $z/R$.  We find that
\begin{equation}
\ddot{x} \approx - x\left[\frac{\partial^2 \Phieff}{\partial R^2}\right]_{R_g}\equiv - \kappa^2(R_g)x
\end{equation}
\noindent
and
\begin{equation}
x(t) \approx X \cos(\kappa t + \psi).
\end{equation}
\noindent
The constant $\kappa$ is called the epicyclic frequency when $\kappa^2>0$.  If $\kappa^2<0$,
the orbit is unstable.

Recalling that in a circular orbit
\begin{equation}
R \Omega^2(R) = \frac{\partial \Phi(R,z=0)}{\partial R}
\end{equation}
\noindent
we can write
\begin{equation}
\kappa^2(R) = \frac{d}{dR}[R\Omega^2(R)] + \frac{3 l_z^2}{R^4} = \frac{1}{R^3}\frac{d}{dR}[(R^2\Omega)^2] = -4B\Omega
\end{equation}
\noindent
where $B$ is Oort's constant.  In the solar neighborhood $B<0$ and $\kappa^2$ is positive and the Solar
orbit is stable.  If $R^2\Omega(R)$, the angular momentum of a circular orbit, increases with $R$ then
all larger circular orbits are stable.  Near a black hole this isn't the case, and the inner
most circular orbit is at $R=6GM/c^2$.

\subsection{Azimuthal Motion in Epicycles}

Since $R$ changes in an epicyclical orbit, the azimuthal velocity must change to
compensate to keep $l_z$ constant.  This means that
\begin{equation}
\dot(\phi) = \frac{l_z}{R^2} = \frac{\Omega(R_g)R_g^2}{(R_g + x)^2} \approx \Omega(R_g)\left(1 - \frac{2x}{R_g}\right).
\end{equation}
\noindent
We can integrate this equation to find that the azimuthal position with time varies as
\begin{equation}
\phi(t) = \phi_0  + \Omega(R_g)t - \frac{1}{R_g}\frac{2\Omega}{\kappa}X \sin(\kappa t + \psi)
\end{equation}
\noindent
The first two terms are circular motion starting from $\phi_0$ with constant angular
frequency $\Omega(R_g)$.  The third term is a harmonic oscillation with the same
frequency $\kappa$ as the $x$ motion but $\pi/2$ out of phase, and
larger by a factor $2\Omega/\kappa$.
The motion is also retrograde!

\subsection{$\kappa$ vs. $\Omega$}

For a point mass $\kappa = \Omega$.  For a uniform sphere $\kappa = 2\Omega$.
The galaxy has $\Omega < \kappa < 2\Omega$, and near the sun $\kappa \approx 1.4 \Omega$.

\subsection{Relative Speed of Nearby Stars}

The relative speeds of stars with $R_g > R_0$ are
\begin{equation}
v_y = R_0[\dot{\phi} - \Omega(R_0)]\approx R_0 \left[ \Omega(R_g) - 2x \frac{\Omega(R_g)}{R_g} - \Omega(R_0)\right]
\end{equation}
\noindent
but since $R_0 = R_g + x$ and dropping terms in $x^2$ we have
\begin{equation}
v_y \approx -x\left[ 2\Omega(R_0) + R_0\left(\frac{d\Omega}{dR}\right)_{R_0}\right] = -\frac{\kappa^2 x}{2\Omega} = 2Bx
\end{equation}
\noindent
Taking the average of stars we can measure, we find
\begin{equation}
\ave{v_y^2} = \left(\frac{\kappa^2}{2\Omega}\right)^2\ave{x^2} = \frac{\kappa^2}{4\Omega^2}\ave{v_x^2}
\end{equation}
\noindent
and since $\kappa<2\Omega$, $\ave{v_y^2}<\ave{v_x^2}$.  For the thin disk, we find that $2\lesssim\ave{v_x^2}/\ave{v_y^2}\lesssim3$.

\end{document}
